### 2023-10-31 doc/15yrsChange.Rmd
update for special issue abstract submission

Luca Morreale

@@

Kristina J. Anderson-Teixeira^1,2^* ( Orcid ID : 0000-0001-8461-9713)

@@

*corresponding author: teixeirak@si.edu; +1 540 635 6546

@@

- temperate deciduous forests play a critical role in the global carbon cycle, accounting for a substantial portion of the global forest C sink
- The dominant view imbued in Earth System Models is that, at least within the eastern US, the biome is likely to remain a C sink for the remainder of the 21st century
- However, these models do not incorporate nuisance species -- i.e., species whose local abundance has been dramatically increased as a result of human activities and that are causing ecological harm. 
- Nuisance species are known to increase tree mortality (non-endemic pests and pathogens) and decrease recruitment (deer, non-endemic plants) throughout the (Eastern US) temperate deciduous biome, yet we don't know net effect on current and future C cycling
- Here, we use 15 years of detailed census data from a large forest dynamics plot with a deer exclosure to understand how nuisance species are affecting C cycling
- we show increased biomass mortality, reduced ∆AGB (only outside deer exclosure?), reduced AGB (only outside deer exclosure?)
- if current trends continue, forest will continue to lose C sequestration capacity/ likely to be C source in future
- thus, C sequestration potential of temperate deciduous forests is overestimated

>>>>

Luca L. Morreale^1,2^*

@@

Kristina J. Anderson-Teixeira^1,2^ ( Orcid ID : 0000-0001-8461-9713)

@@

*corresponding author: MorrealeLL@si.edu; +1 607 227 1350

@@

Temperate deciduous forests play a critical role in the global carbon (C) cycle, accounting for a substantial portion of the global forest C sink. The dominant view imbued in Earth System Models is that, at least within the eastern United States, the biome is likely to remain a C sink throughout the 21st century. However, these models do not incorporate nuisance species -- *i.e.*, species whose local abundance has increased as a result of human activities and that are causing ecological harm. Nuisance species are known to increase tree mortality (*e.g.*, non-endemic pests and pathogens) and decrease recruitment (*e.g.*, deer and non-endemic plants) throughout the temperate deciduous biome, yet the net effect on current and future C cycling is unknown. Here, we use 15 years of detailed census data from a large (25.6 ha) forest dynamics plot in a Mid-Atlantic temperate forest, including a 4-ha deer exclosure, to understand how nuisance species are affecting C cycling. We show increased biomass mortality, a net reduction in aboveground C storage, and decreased abundance of canopy tree species in the understory. Under current trends this forest will continue to lose C sequestration capacity, indicating that the C sequestration of temperate deciduous forests is likely overestimated.
